Donkeys to Monoids: A Computational Framework for Dynamic Semantics

### 2013-05-19 paper.tex
Fixed DPL part

In the previous chapter I showed how first-order logic lacks certain compositional properties. A possible solution to this problem is to use a so called `update logic' (a logic with update semantics). Update semantics are logics inspired on the semantics of programming languages where a variable, once introduced, will stick around for a period not know at the time of it's introduction.

@@

The logic cleverly uses these advances in logical understanding to create an update logic in a pragmatic way, without any external constructs and a minimum of new syntax, while still getting all the benefits of compositionality etc.

To model a simple sentence `A man comes in. He sees a donkey. He smiles.', we write:
%
\begin{align}
&(\exists x \cdot man(x) \cdot comes\_in(x)) \nonumber\\
\cdot\ &(\exists y \cdot donkey(y) \cdot sees(x,y)) \nonumber\\
\cdot\ &(smiles(x)) \nonumber
\end{align}

@@

Too understand the semantics of the above, let's start by only considering a dynamic propositional logic with three propositions: $P$, $Q$ and $R$. We'll write $p$ to indicate a predicate is true and $\overline{p}$ if it's false. We talked about earlier, that the update semantics modeled `growth of information over time', so we'll start with case that every $2^3$ state is possible:
%
\begin{align}
\{\overline{PQR}, \overline{PQ}R, \overline{P}Q\overline{R}, \overline{P}QR, P\overline{QR}, P\overline{Q}R, PQ\overline{R}, PQR\} \label{dpl_all_states}
\end{align}

Now we'll consider a statement in our predicate logic as a relation between possible states. For example $P\overline{Q}R[P\cdot Q]P\overline{Q}R$ is false, while $P\overline{Q}R[P\cup Q]PQR$ is true. If we take the image of \eqref{dpl_all_states} under $P\cdot\neg Q$ we get
%
\begin{align}
\{P\overline{QR}, P\overline{Q}R\} \label{dpl_PnQ_states}
\end{align}

If we then take the image of \eqref{dpl_PnQ_states} under $\overline{P}$ we end up with the empty set of states. That is there is no possible assignment satisfying our formula, and we consider it a falsum.

The most interesting question is how we work with quantifiers. Which on earth relation could $\exists S$ be? We are going to take a hint from computer programming and consider the variable declaration: $S:=\bot$. We want to write this in DPL as $\exists S \cdot \neg S$, such that e.g. $P\overline{Q}[\exists S \cdot \neg S]P\overline{QS}$ is true. Now since $\cdot\neg S$ acts as a filter, removing all assignments that doesn't have $\overline{S}$, it must be that $\exists S$ is a `random reset, introducing $S$ and allowing it every value, such that $P\overline{Q}[\exists S]P\overline{QS}$ and $P\overline{Q}[\exists S]P\overline{Q}S$ are both true. The image of \eqref{dpl_all_states} under $\exists S$ has 16 possible states.

Let's have a look at the model theoretic definition of the semantics. Visser builds this up in two steps: First a relational algebra and then DPL on top of it. But since DPL is basically identical to the relational algebra, we are going to try and do it in one step.

@@

We are further going to define:
%
\begin{align}
Prop_\Sigma &= \{P(x_1,\dots,x_n) \mid P \in Pred,\ Ar(P)=n,\ x_1,\dots,x_n\in Var\}\\
Reset_\Sigma &= \{\exists x \mid x\in Var\}\\
Atom_\Sigma &= Var \cup Prop_\Sigma \cup Reset_\Sigma
\end{align}

A model $\mathcal{M}$ of a signature $\Sigma$ is a tuple $\langle\mathcal{D},\mathcal{I}\rangle$ where $\mathcal{D}$ is a domain which we will always set as $\{\top,\bot\}$, and $\mathcal{I}$ is the interpretation function $Prop_\Sigma\to\mathcal{D}$.

@@

We are also going to say that an assignment can satisfy a formula, defined as:
\begin{align}
\alpha\models\psi :\equiv \exists\beta(\alpha[\psi]\beta)
\end{align}

Let $[\cdot]$ be the interpretation function from syntax to relation. Then we define the semantics of our logic pointwise for all assignments $\alpha$ and $\beta$:
\begin{align}
\alpha[\psi_1\cdot\psi_2]\beta :\equiv&\ \exists\gamma (\alpha[\psi_1]\gamma \wedge \gamma[\psi_2]\beta) \label{sem_and} \\
\alpha[\psi_1\cup\psi_2]\beta :\equiv&\ \alpha[\psi_1]\beta \vee \alpha[\psi_2]\beta \label{sem_union} \\
\alpha[P(x_1,\dots,x_n)]\beta :\equiv&\ \alpha = \beta \wedge \mathcal{I}(P(\alpha_{x_1},\dots,\alpha_{x_n})) \label{sem_pred} \\
\alpha[\neg\psi]\beta :\equiv&\ \alpha = \beta \wedge \alpha\not\models[\psi] \label{sem_neg}\\
\alpha[\exists x]\beta :\equiv&\ \forall\omega (\omega \neq x \rightarrow \alpha_{\omega} = \beta_{\omega}) \label{sem_exists} \\
\intertext{Using those basic axioms we can derive for our built in propositions:}
\alpha[\bot]\beta \leftrightarrow&\ \alpha = \beta \wedge \bot \nonumber\\
                  \leftrightarrow&\ \bot \label{sem_bot}\\
\alpha[\top]\beta \leftrightarrow&\ \alpha = \beta \wedge \top \nonumber\\
                  \leftrightarrow&\ \alpha = \beta \label{sem_top}\\
\intertext{Taking $\psi_1\rightarrow\psi_2$ to mean $\neg(\psi_1\wedge\neg\psi_2)$ we also get the following neat equation, originally by Groenendijk and Stokhof\cite{groenendijk1991dynamic}:}
\alpha[\psi_1\rightarrow\psi_2]\beta
 :\equiv&\ \alpha[\neg(\psi_1\wedge\neg\psi_2)]\beta \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \alpha\not\models[\psi_1\wedge\neg\psi_2] \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \neg\exists\gamma(\alpha[\psi_1\wedge\neg\psi_2]\gamma) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \neg\exists\gamma(\exists\delta(\alpha[\psi_1]\delta\wedge\delta[\neg\psi_2]\gamma)) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \neg\exists\gamma(\exists\delta(\alpha[\psi_1]\delta\wedge\delta=\gamma\wedge\delta\not\models[\psi_2])) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \forall\gamma(\forall\delta(\delta=\gamma\rightarrow(\alpha[\psi_1]\delta\rightarrow\delta\models[\psi_2]))) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \forall\gamma(\alpha[\psi_1]\gamma\rightarrow\gamma\models[\psi_2]) \label{sem_impl}
\end{align}

In my implementation section, I describe how the above can be efficiently implemented in Haskell using an isomorphism between relations $\mathcal{P}(A\times B)$ and functions $A\to\mathcal{P}(B)$. I also show how the kleiski combinator takes the role of $\cdot$.

However let's first look at some of the implications of the above definitions. For example, how can we intuitively understand the DPL notion of $\rightarrow$? Well, first notice that many of the equations start with $\alpha=\beta\wedge\dots$. These equations we can think of intuitively as having a closed scope. We saw earlier how DRT nicely determined that the `it' in `A farmer doesn't own a donkey. He beats it' couldn't refer to `a donkey' because the negation didn't let it's variable out. In DPL we get the same effect. Because an assignment must be the same before and after a negated clause, all new variables introduced inside the clause have disappeared.

Of course this `static negation' might not always be the right thing. Consider the sentences `It is not that case that a farmer doesn't own a donkey. He beats it'. This should be a good sentence, but we can't handle it. In the next chapter we will look at how this might be handled.

@@

An interesting property of DPL is that it is isomorphic with classical logic. Any PL formula corresponds to one in DPL and vice versa. This gives an alternative way to define and understand the semantics of DPL, and it relaxes us that nothing `weird' is going on. We might also be a bit disappointed that after all this, we don't possess any extra power, but let's forget about that for now, and see how the conversion works.

Let $\phi$ be a formula in negation normal form, or something. Hmm, come back here once the disjunction chapter is done.
\begin{align}
&[\exists x(\phi)] \equiv \neg(\neg(\exists x \cdot [\phi]))\\
&[\forall x(\phi)] \equiv \exists x \rightarrow [\phi]\\
&[\phi_1\wedge\phi_2] \equiv [\phi_1]\cdot[\phi_2]\\
&[\phi_1\vee\phi_2] \equiv \dots\\
&[\neg(\exists x)] \equiv \dots\\
&[\neg(\phi_1\wedge\phi_2)] \equiv \neg[\phi_1]\cdot\neg[\phi_2]\\
&[\neg(\phi_1\vee\phi_2)] \equiv \dots
\end{align}

Similarly we may create PL formulas with the same meaning as any DPL formula. But perhaps we should be a bit concerned on how to interpret the meaning of a PL formula and a DPL formula being equivalent. After all DPL formulas are relations where PL formulas a propositions. In the above it was all easy, but let's think again about what we are doing.

Remember that we said an assignment could model a DPL relation. If we consider the truth value of a DPL relation to be whether there exists any relation that models it $\exists\alpha(\alpha\models\psi)$, we can use a trick. The trick is to think of the weakest precondition such an assignment must satisfy, and model that in PL. In fact the precondition is not just for our assignment, but for the entire model. For example to model $\exists x\cdot farmer(x)$ the starting assignment might well be empty ($\bot$), but our model must contain an entity for which $farmer$ is true.

To smoothen the following derivations, we will introduce a special syntax `$\langle\psi\rangle\phi$' where $\psi$ is a DRA formula, $\phi$ is a PL formula and the whole thing is a PL formula. For some assignment $\alpha$, we define `$\alpha\models\langle\psi\rangle\phi$' to mean $\exists\beta(\alpha[\psi]\beta \wedge \beta\models\phi)$.
%
\begin{align}
\alpha\models\langle\exists x\rangle\phi
 & \leftrightarrow \alpha\models\exists x (\phi) \label{conv_exists}\\
\alpha\models\langle\psi_1\cdot\psi_2\rangle\phi
 & \leftrightarrow \alpha\models\langle\psi_1\rangle\langle\psi_2\rangle\phi \label{conv_and}\\
\alpha\models\langle\psi_1 \cup \psi_2\rangle\phi
 & \leftrightarrow \alpha\models\langle\psi_1\rangle\phi \vee \langle\psi_2\rangle\phi \label{conv_or}\\
 \alpha\models\langle\psi_1 \vee \psi_2\rangle\phi
\alpha\models\langle P(x_1,\dots,x_n)\rangle\phi
 & \leftrightarrow \alpha\models P(x_1,\dots,x_n) \wedge \phi \label{conv_pred}\\
\alpha\models\langle\neg\psi\rangle\phi
 & \leftrightarrow \alpha\models \neg\langle\psi\rangle\top \wedge \phi \label{conv_neg}
\end{align}

See the appendix for my full derivations of every equation. Also notice that $\alpha\models\langle\bot\rangle\phi\leftrightarrow a\models\bot$ and $\alpha\models\langle\top\rangle\phi\leftrightarrow a\models\phi$ follow trivially from \eqref{conv_pred}. Equations for `convenience' operators such as $\rightarrow$ and $\vee$ can be derived too:
\begin{align}
\langle\psi_1\vee\psi_2\rangle\phi
& \leftrightarrow \langle\neg(\neg\psi_1\cdot\neg\psi_2)\rangle\phi \nonumber\\
& \leftrightarrow \neg\langle\neg\psi_1\cdot\neg\psi_2\rangle\top \wedge\phi \nonumber\\
& \leftrightarrow \neg\langle\neg\psi_1\rangle\langle\neg\psi_2\rangle\top \wedge\phi \nonumber\\
& \leftrightarrow \neg\langle\neg\psi_1\rangle(\neg\langle\psi_2\rangle\top) \wedge\phi \nonumber\\
& \leftrightarrow \neg(\neg\langle\psi_1\rangle\top \wedge \neg\langle\psi_2\rangle\top) \wedge\phi \nonumber\\
& \leftrightarrow (\langle\psi_1\rangle\top \vee \langle\psi_2\rangle\top) \wedge\phi \label{conv_vee}
\end{align}

In the implementation we will use the above equations to recursively convert our DPL formulas to PL. The $\langle\rangle$ notation easily lends itself to calculation by a stack manipulation function.

The rules will look like:
\begin{haskell}
addDpl :: Prop -> [DPL] -> Prop
addDpl p ((DNot d):ds) = addDpl (And (Not (addDpl T [d])) p) ds
\end{haskell}

@@

\subsubsection{Examples}

@@

A few examples of sentences that my framework already supports, but which fail due to short comings in the Stanford toolkit are `Every dog sees a cat. It chases it.' where the anaphora doesn't get resolved, and `Buffalo buffalo Buffalo buffalo buffalo buffalo Buffalo buffalo' that manages to slightly confuse the parser.

@@

Finally I wish to thanks to The Donkey Sanctuary\cite{donkey2013sanctuary} for taking donkey's welfare seriously and rescuing neglected donkeys in the UK and all over the world.

@@

\subsection{Derivation of DPL Preconditions}
Remember that we have defined, for assignments $\alpha$ and $\beta$, DPL formula $\psi$ and FOL formula $\phi$: $\alpha\models\langle\psi\rangle\phi$ to mean $\exists\beta(\alpha\llbracket\psi\rrbracket\beta\wedge\beta\models\phi)$. $\alpha\models\phi$ is just the usual $\models$ from FOL.

@@

 & \leftrightarrow \alpha\models\exists x (\phi) \\

@@

 & \leftrightarrow \alpha\models\langle\psi_1\rangle\langle\psi_2\rangle\phi

@@

 & \leftrightarrow \alpha\models\langle\psi_1\rangle\phi \vee \langle\psi_2\rangle\phi

@@

 & \leftrightarrow \alpha\models P(x_1,\dots,x_n) \wedge \phi

@@

 & \leftrightarrow \alpha\models \neg\langle\psi\rangle\top \wedge \phi

@@

\intertext{Because $\psi$ is this way `evaluated on $\top$ only', we know that no variables introduced in $\psi$ can affect $\phi$.}\nonumber

@@

This sentence suggests that we might want to define a disjunction operator with a linear flow of variables: $\psi_1\to\psi_2\to\psi_3$ instead of $\psi_1\to\psi_3\text{ and }\psi_2\to\psi_3$. However there is a problem: We currently define `false' as the absence of possible assignments. Hence if $\psi_1$ is false, there is no way we can define disjunction to push it's variables into $\psi_2$. If any variables were left undefined, $\psi_1$ wouldn't be false.

There is a cure to this problem. When we extend our logic with more power in the monoid chapter, we get a second chance. However Let's first have a look at an obvious alternative definition for disjunction considered by Groenendijk\cite{groenendijk1991dynamic}. We'll call it `static disjunction':
\begin{align}
\alpha[\psi_1\vee\psi_2]\beta
 :\equiv&\ \alpha[\neg(\neg\psi_1\wedge\neg\psi_2)]\beta \nonumber\\
 \leftrightarrow&\ \alpha[\neg\psi_1\rightarrow\psi_2]\beta \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \forall\gamma(\alpha[\neg\psi_1]\gamma\rightarrow\gamma\models[\psi_2]) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \forall\gamma((\alpha=\gamma\wedge\gamma\not\models\psi_1)\rightarrow\gamma\models[\psi_2]) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \forall\gamma(\alpha=\gamma\rightarrow(\gamma\models\psi_1\vee\gamma\models\psi_2)) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge (\alpha\models\psi_1\vee\alpha\models\psi_2) \label{sem_or}
\end{align}

The idea to this follows natural from De Morgan's laws of classical logic. Clearly the semantics are different from the ones we defined for $\cup$. However not in a good way. This static disjunction doesn't allow any variables to escape from inside it's reach.

@@

\subsection{Source Code}

>>>>

In the previous chapter I showed how first-order logic lacks certain compositional properties. A possible solution to this problem is to use a so called `update logic' (a logic with update semantics). Update semantics are logics inspired on the semantics of programming languages where a variable, once introduced, will stick around for a period not know at the time of its introduction.

@@

DPL cleverly uses this basic logical research to create a very simple, self-contained logic, while still getting all the benefits of compositionality etc. To model a simple sentence `A man comes in. He sees a donkey. He smiles.', we write:
%
\begin{align}
&(\exists x \cdot man(x) \cdot comes\_in(x)) \nonumber\\
\cdot\ &(\exists y \cdot donkey(y) \cdot sees(x,y)) \nonumber\\
\cdot\ &(smiles(x)) \nonumber
\end{align}

@@

Too understand the semantics of the above, let's start by considering a simple example with two variables ($a$ and $b$) and a domain with three entities ($0$, $1$ and $2$). I said earlier, that the update semantics modeled `growth of information over time', so we'll start with the case that every $3^2$ assignments are possible:
%
\begin{align}
\{&\langle a\mapsto0,b\mapsto0\rangle, \langle a\mapsto0,b\mapsto1\rangle, \langle a\mapsto0,b\mapsto2\rangle,\nonumber\\
&\langle a\mapsto1,b\mapsto0\rangle, \langle a\mapsto1,b\mapsto1\rangle, \langle a\mapsto1,b\mapsto2\rangle,\nonumber\\
&\langle a\mapsto2,b\mapsto0\rangle, \langle a\mapsto2,b\mapsto1\rangle, \langle a\mapsto2,b\mapsto2\rangle\}\label{dpl_all_states}
\end{align}

A DPL formula is defined as a relation between assignments, and $\llbracket\cdot\rrbracket$ is the function that takes a formula to its relation. For example $\langle a\mapsto1,b\mapsto1\rangle\llbracket a=1\rrbracket\langle a\mapsto1,b\mapsto1\rangle$ is true because the $a=1$ `program' executes successfully. We also require that simple tests like this one doesn't change the state, so $\langle a\mapsto 1,b\mapsto2\rangle$ would not have been a valid assignment on the right hand side.

Just like with programs, we can combine multiple formulas into bigger ones: $\langle a\mapsto0,b\mapsto2\rangle\llbracket a=0\cdot b=a\rrbracket\psi$ is false for all $\psi$, and $\langle a\mapsto2,b\mapsto1\rangle\llbracket a=0\cup b<2\rrbracket\langle a\mapsto2,b\mapsto1\rangle$ is true. If we take the image of \eqref{dpl_all_states} under $a=0\cdot\neg (b=2)$ we get
%
\begin{align}
\{&\langle a\mapsto0,b\mapsto0\rangle, \langle a\mapsto0,b\mapsto1\rangle\} \label{dpl_PnQ_states}
\end{align}

If we take the image of \eqref{dpl_PnQ_states} under $a=1$, we end up with the empty set of states. That is because there are no possible assignments satisfying our formula. Hence we consider $a=0\cdot\neg (b=2)\cdot a=1$ a falsum, also defined as $\bot$.

So why do we define all of this using relations, and not just functions from states to truth values? The reason is that we want to introduce a special relation $\exists x$ (for some variable name $x$). We want it to analogous to $\exists x(...)$ from first-order logic. It's not immediately obvious what the correct definition should be.

The trick is to consider the conjunctive DPL relation $\exists x\cdot x=1$ - we want this to work such that $\langle a\mapsto0\rangle\llbracket\exists x\cdot x=1\rrbracket\langle a\mapsto0,x\mapsto1\rangle$. Now considering $x=1$ is a filter, it must be that $\exists x$ is a relation that allows any value for $x$ on its second argument. For example we see that the image of \eqref{dpl_all_states} under $\exists x$ has 27 possible states.

Let's have a look at the model theoretic definition of the semantics. This is sometimes done in two steps: First defining a relational algebra and then builid DPL on top of it. However DPL is basically identical to the relational algebra, so we are going to try and do everything in one step:

@@

We are further going to define:
%
\begin{align}
Prop_\Sigma &= \{P(x_1,\dots,x_n) \mid P \in Pred,\ Ar(P)=n,\ x_1,\dots,x_n\in Var\}\\
Reset_\Sigma &= \{\exists x \mid x\in Var\}\\
Atom_\Sigma &= Prop_\Sigma \cup Reset_\Sigma
\end{align}

@@

A model $\mathcal{M}$ of a signature $\Sigma$ is a tuple $\langle\mathcal{D},\mathcal{I}\rangle$. The model is very similar to first-order logic in that $\mathcal{D}$ is the domain, a set of entities, and $\mathcal{I}$ is the interpretation function $Prop_\Sigma\to(Asgmt\to\mathbb{B})$, ($Asgmt$ being the set of possible assignments, $\mathbb{B}$ the set of boolean values).

Just like in first-order logic we want to denote whether a formula is true or false, and just like in first-order logic we do this by saying that an assignment can satisfy a formula, defined as below. If a formula is satisfied by all assignments (equivalent to being satisfied by the empty assignment), it is a tautology.
\begin{align}
\alpha\models\psi :\equiv \exists\beta(\alpha\llbracket\psi\rrbracket\beta)
\end{align}

We then define our relations in the set theoretic way, as sets of pairs:
\begin{align}
\llbracket\psi_1\cdot\psi_2\rrbracket :\equiv&\ \{\langle\alpha,\beta\rangle \in Asgmt^2 \mid \exists\gamma(\langle\alpha,\gamma\rangle\in\llbracket\psi_1\rrbracket\wedge\langle\gamma,\beta\rangle\in\llbracket\psi_2\rrbracket)\} \label{sem_set_and} \\
\llbracket P(x_1,\dots,x_n)\rrbracket :\equiv&\ \{\langle\alpha,\alpha\rangle \in Asgmt^2 \mid \mathcal{I}(P)(\alpha)\} \label{sem_set_pred}
\end{align}
%
Or we can define them more succinctly, pointwise for all assignments $\alpha$, $\beta$:
\begin{align}
\alpha\llbracket \psi_1\cdot\psi_2\rrbracket \beta :\equiv&\ \exists\gamma (\alpha\llbracket \psi_1\rrbracket \gamma \wedge \gamma\llbracket \psi_2\rrbracket \beta) \label{sem_and} \\
\alpha\llbracket \psi_1\cup\psi_2\rrbracket \beta :\equiv&\ \alpha\llbracket \psi_1\rrbracket \beta \vee \alpha\llbracket \psi_2\rrbracket \beta \label{sem_union} \\
\alpha\llbracket P(x_1,\dots,x_n)\rrbracket \beta :\equiv&\ \alpha = \beta \wedge \mathcal{I}(P)(\alpha) \label{sem_pred} \\
\alpha\llbracket \neg\psi\rrbracket \beta :\equiv&\ \alpha = \beta \wedge \alpha\not\models\llbracket \psi\rrbracket  \label{sem_neg}\\
\alpha\llbracket \exists x\rrbracket \beta :\equiv&\ \forall\omega (\omega \neq x \rightarrow \alpha_{\omega} = \beta_{\omega}) \label{sem_exists}
\end{align}
%
Using those basic axioms we can derive $\top$ and $\bot$ given our usual interpretations:
\begin{align}
\alpha\llbracket \bot\rrbracket \beta \equiv&\ \alpha = \beta \wedge \mathcal{I}(\bot)(\alpha) \equiv \bot \label{sem_bot}\\
\alpha\llbracket \top\rrbracket \beta \equiv&\ \alpha = \beta \wedge \mathcal{I}(\top)(\alpha) \equiv \alpha = \beta \label{sem_top}\\
\end{align}
%
Taking $\psi_1\rightarrow\psi_2$ to mean $\neg(\psi_1\wedge\neg\psi_2)$ we also get the following neat equation, originally by Groenendijk and Stokhof\cite{groenendijk1991dynamic}
\begin{align}
\alpha\llbracket \psi_1\rightarrow\psi_2\rrbracket \beta
 :\equiv&\ \alpha\llbracket \neg(\psi_1\wedge\neg\psi_2)\rrbracket \beta \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \alpha\not\models\llbracket \psi_1\wedge\neg\psi_2\rrbracket  \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \neg\exists\gamma(\alpha\llbracket \psi_1\wedge\neg\psi_2\rrbracket \gamma) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \neg\exists\gamma(\exists\delta(\alpha\llbracket \psi_1\rrbracket \delta\wedge\delta\llbracket \neg\psi_2\rrbracket \gamma)) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \neg\exists\gamma(\exists\delta(\alpha\llbracket \psi_1\rrbracket \delta\wedge\delta=\gamma\wedge\delta\not\models\llbracket \psi_2\rrbracket )) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \forall\gamma(\forall\delta(\delta=\gamma\rightarrow(\alpha\llbracket \psi_1\rrbracket \delta\rightarrow\delta\models\llbracket \psi_2\rrbracket ))) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \forall\gamma(\alpha\llbracket \psi_1\rrbracket \gamma\rightarrow\gamma\models\llbracket \psi_2\rrbracket ) \label{sem_impl}
\end{align}

In my implementation section, I describe how the above can be efficiently implemented in Haskell using an isomorphism between relations $\mathcal{P}(A\times B)$ and functions $A\to\mathcal{P}(B)$. I also show how the Kleisli combinator takes the role of $\cdot$.

However let's first look at some of the implications of the above definitions. For example, how can we intuitively understand the DPL notion of $\rightarrow$? Well, first notice that many of the equations start with $\alpha=\beta\wedge\dots$. These equations we can think of intuitively as having a closed scope. We saw earlier how DRT nicely determined that the `it' in `Some farmer owns no donkey. He beats it' couldn't refer to `a donkey' because the negation didn't let its variable out. In DPL we get the same effect. Because an assignment must be the same before and after a negated clause, all new variables introduced inside the clause have disappeared.

Of course this `static negation' might not always be the right thing. Consider the sentences `It is not that case that a farmer doesn't own a donkey. He beats it'. This should be a good sentence, but we can't handle it. In section~\ref{sec:problem_disj} I will look at how this might be handled.

@@

An interesting property of DPL is that it is isomorphic with classical logic. Any PL formula corresponds to one in DPL and vice versa. This gives an alternative way to define and understand the semantics of DPL, and it relaxes us that nothing `weird' is going on. We might also be a bit disappointed that after all this work, we don't possess any extra power, but let's forget about that for now, and see how the conversion works.

We can define a conversion function $\llbracket\cdot\rrbracket$ from FOL to DPL very easily. The only trick is that we need a way in DPL to create the closed scopes used in FOL. $\neg(\neg(\cdot))$ does nicely:
\begin{align}
\llbracket P(x_1,\dots,x_n)\rrbracket &\equiv P(x_1,\dots,x_n)\\
\llbracket \neg(\phi)\rrbracket &\equiv \neg\llbracket \phi\rrbracket\\
\llbracket \phi_1\wedge\phi_2\rrbracket &\equiv \llbracket \phi_1\rrbracket\cdot\llbracket \phi_2\rrbracket\\
\llbracket \phi_1\vee\phi_2\rrbracket &\equiv \llbracket \phi_1\rrbracket\cup\llbracket \phi_2\rrbracket\\
\llbracket \exists x(\phi)\rrbracket &\equiv \neg(\neg(\exists x \cdot \llbracket \phi\rrbracket))\\
\llbracket \forall x(\phi)\rrbracket &\equiv \exists x \rightarrow \llbracket \phi\rrbracket
\end{align}

Similarly we can create a function from DPL to FOL, but perhaps we should be a bit concerned on how to interpret the meaning of an FOL formula and a DPL formula being equivalent. After all DPL formulas are relations where PL formulas are propositions. In the above it was all easy, but let's think again about what we are doing.

We are going to use our definition of $\alpha\models\phi$ for DPL formulas. An FOL formula $\phi$ and a DPL formula $\psi$ are defined to be equivalent exactly when $\forall\alpha(\alpha\models\phi\leftrightarrow\alpha\models\psi)$. Perhaps the best way to see why this is a good definition (other than its simplicity) is to think programs and weakest preconditions. Every initial state for the `computation' $farmer(y)\cdot\exists x\cdot x=y$ must model $farmer(y)\wedge\exists x(x=y)$ in order to succeed.

To smoothen the following derivations, we will define the function $\langle\psi\rangle\phi:DPL\times FOL\to FOL$ by
\begin{align}
\alpha\models\langle\psi\rangle\phi\ \leftrightarrow\ \exists\beta(\alpha\llbracket\psi\rrbracket\beta \wedge \beta\models\phi)\label{foldpl}
\end{align}
Notice from this definition that $\langle\psi\rangle\top$ is the conversion function from DPL to FOL.
Using definition \eqref{foldpl} and formulas \eqref{sem_and} to \eqref{sem_exists} we can derive the following equivalences. See appendix~\ref{sec:derivation_dpl} for my full derivations.
%
\begin{align}
\langle\exists x\rangle\phi & \equiv \exists x (\phi) \label{conv_exists}\\
\langle\psi_1\cdot\psi_2\rangle\phi & \equiv \langle\psi_1\rangle\langle\psi_2\rangle\phi \label{conv_and}\\
\langle\psi_1 \cup \psi_2\rangle\phi & \equiv \langle\psi_1\rangle\phi \vee \langle\psi_2\rangle\phi \label{conv_or}\\
\langle P(x_1,\dots,x_n)\rangle\phi & \equiv  P(x_1,\dots,x_n) \wedge \phi \label{conv_pred}\\
\langle\neg\psi\rangle\phi & \equiv  \neg\langle\psi\rangle\top \wedge \phi \label{conv_neg}\\
\langle\psi_1\vee\psi_2\rangle\phi & \equiv (\langle\psi_1\rangle\top \vee \langle\psi_2\rangle\top) \wedge\phi \label{conv_vee}
\end{align}

The simple predicates $\langle\bot\rangle\phi\equiv\bot$ and $\langle\top\rangle\phi\equiv\phi$ follow trivially from \eqref{conv_pred}. Using the first 5 equations, equivalences for derives operators such as $\rightarrow$ and $\vee$ can be derived syntactically as also shown in appendix~\ref{sec:derivation_dpl}.

An alternative way to the $\langle\psi\rangle\phi$ function for looking at the above equations is to think of the $\langle\cdot\rangle$s as elements on a stack. This is indeed how I have implemented the DPL to FOL routine in appendix~\ref{sec:source_code}.

@@

\subsubsection{A Bunch of Examples}

@@

A few examples of sentences that my framework already supports, but which fail due to shortcomings in the Stanford toolkit are `Every dog sees a cat. It chases it.' where the anaphora doesn't get resolved, and `Buffalo buffalo Buffalo buffalo buffalo buffalo Buffalo buffalo' that manages to slightly confuse the parser.

@@

Finally I wish to thank The Donkey Sanctuary\cite{donkey2013sanctuary} for taking donkey's welfare seriously and rescuing neglected donkeys in the UK and all over the world.

@@

This sentence suggests that we might want to define a disjunction operator with a linear flow of variables: $\psi_1\to\psi_2\to\psi_3$ instead of $\psi_1\to\psi_3\text{ and }\psi_2\to\psi_3$. However there is a problem: We currently define `false' as the absence of possible assignments. Hence if $\psi_1$ is false, there is no way we can define disjunction to push its variables into $\psi_2$. If any variables were left undefined, $\psi_1$ wouldn't be false.

There is a cure to this problem. When we extend our logic with more power in the monoid chapter, we get a second chance. However let's first have a look at an obvious alternative definition for disjunction considered by Groenendijk\cite{groenendijk1991dynamic}. We'll call it `static disjunction':
\begin{align}
\alpha[\psi_1\vee\psi_2]\beta
 :\equiv&\ \alpha[\neg(\neg\psi_1\wedge\neg\psi_2)]\beta \label{sem_or_def}\\
 \leftrightarrow&\ \alpha[\neg\psi_1\rightarrow\psi_2]\beta \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \forall\gamma(\alpha[\neg\psi_1]\gamma\rightarrow\gamma\models[\psi_2]) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \forall\gamma((\alpha=\gamma\wedge\gamma\not\models\psi_1)\rightarrow\gamma\models[\psi_2]) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \forall\gamma(\alpha=\gamma\rightarrow(\gamma\models\psi_1\vee\gamma\models\psi_2)) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge (\alpha\models\psi_1\vee\alpha\models\psi_2) \label{sem_or}
\end{align}

The idea to this follows naturally from De Morgan's laws of classical logic. Clearly the semantics are different from the ones we defined for $\cup$. However not in a good way. This static disjunction doesn't allow any variables to escape from inside its reach.

\subsection{Derivation of DPL Preconditions}\label{sec:derivation_dpl}
Remember that we have defined, for assignments $\alpha$ and $\beta$, DPL formula $\psi$ and FOL formula $\phi$: $\alpha\models\langle\psi\rangle\phi$ to mean $\exists\beta(\alpha\llbracket\psi\rrbracket\beta\wedge\beta\models\phi)$. $\alpha\models\phi$ is just the usual $\models$ from FOL.

@@

 & \leftrightarrow \alpha\models\exists x (\phi) \tag{$\ref{conv_exists}$} \\

@@

 & \leftrightarrow \alpha\models\langle\psi_1\rangle\langle\psi_2\rangle\phi \tag{$\ref{conv_and}$}

@@

 & \leftrightarrow \alpha\models\langle\psi_1\rangle\phi \vee \langle\psi_2\rangle\phi \tag{$\ref{conv_or}$}

@@

 & \leftrightarrow \alpha\models P(x_1,\dots,x_n) \wedge \phi \tag{$\ref{conv_pred}$}

@@

 & \leftrightarrow \alpha\models \neg\langle\psi\rangle\top \wedge \phi \tag{$\ref{conv_neg}$}

@@

\intertext{Because $\psi$ is this way `evaluated on $\top$ only', we know that no variables introduced in $\psi$ can affect $\phi$.
The derivation of $\vee$ can be done from the previous results alone:}
\langle\psi_1\vee\psi_2\rangle\phi
& \equiv \langle\neg(\neg\psi_1\cdot\neg\psi_2)\rangle\phi & \text{by \eqref{sem_or_def}}\nonumber\\
& \equiv \neg\langle\neg\psi_1\cdot\neg\psi_2\rangle\top \wedge\phi & \text{by \eqref{conv_neg}}\nonumber\\
& \equiv \neg\langle\neg\psi_1\rangle\langle\neg\psi_2\rangle\top \wedge\phi & \text{by \eqref{conv_and}}\nonumber\\
& \equiv \neg\langle\neg\psi_1\rangle(\neg\langle\psi_2\rangle\top) \wedge\phi & \text{by \eqref{conv_neg}}\nonumber\\
& \equiv \neg(\neg\langle\psi_1\rangle\top \wedge \neg\langle\psi_2\rangle\top) \wedge\phi & \text{by \eqref{conv_neg}}\nonumber\\
& \equiv (\langle\psi_1\rangle\top \vee \langle\psi_2\rangle\top) \wedge\phi &\text{by \eqref{sem_or_def}} \tag{$\ref{conv_vee}$}

@@

@@

\subsection{Source Code}\label{sec:source_code}
